Extending Graphical Representations for Compact Closed Categories with Applications to Symbolic Quantum Computation

### 2008-05-02 papers/reasoning-graphically/paper.tex
added some explaination about the connections to QC more generally. Need to add the references to the bibtex!
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1468 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

While compact closed categories provide a suitable setting for
reasoning about quantum computation \cite{AbrCoe:CatSemQuant:2004},
freely generated structure will not suffice:  we need additional
equations.  These equations will be expressed as rewrites rules for
graphs.  In this section we will describe a set of generators and equations
used to reason about quantum computation, and show how some of its formal
properties lead to particular issues for the rewriting machinery.

@@

>>>>

A substantial strand of work in quantum computation is
the development of high-level models of quantum processes based on
compact closed  categories.  In this formalism, initiated in
\cite{AbrCoe:CatSemQuant:2004}, quantum processes---such as quantum
logic gates, or the measurement of a qubit---correspond to arrows
in a category, while the different quantum data types, usually just
arrays of qubits, are the objects.  The graphical presentation
described in the the preceding section thus provides a very expressive
notation for representing quantum processes.

While compact closed categories provide a suitable setting for
reasoning about quantum computation,
freely generated structure will not suffice:  we need additional
equations.  These equations will be expressed as rewrites rules for
graphs.  In this section we will describe a set of generators and equations
used to reason about quantum computation, and show how some of its formal
properties lead to particular issues for the rewriting machinery.

@@

Space does not permit a full exposition of this system, however we
can outline the basic intuitions behind these generators.  Each edge
in the picture represents a qubit, although several edges may
represent the same physical qubit at different times.  An edge may
even represent a ``virtual'' qubit which stands for a correlation
between different parts of the  system.  The maps
$\delta_Z$ and $\epsilon_Z$ represent quantum operations which
respectively copy and delete the eigenstates
of the  Pauli $Z$ operator.\footnote{Uniform copying operations are
  forbidden by the no-cloning theorem \cite{no-cloning}, but such
  operations are possible if we demand only the eigenstates of some
  self-adjoint operator to be copied. Other states will not not
  copied.  The same remarks hold true for erasing.}  
Notice that $\delta_Z$ has one edge
in its domain representing the qubit to copied, and two edges in its
codomain for the two copies it produces.  Similarly, $\epsilon_Z$ has
one qubit as input and no outputs.
The adjoints $\delta^\dag_Z$ and  $\epsilon^\dag_Z$ correspond to an
operation known as \emph{fusion}, and to the operation of preparing a
fresh qubit in a certain state. The $\alpha_Z$ corresponds to phase
shift of angle $\alpha$ in the $Z$ direction.  The family of maps
indexed by $X$ are defined in exactly the same way, but relative to
the Pauli $X$ rather than the $Z$.  All quantum operations may be
defined by combining these simple operations---which are essentially
classical---on two complementary observables.

\begin{remark}
  We emphasise that this is a mathematical notation is for
  representing quantum processes, unlike the so-called graph states
  used in measurement based quantum computation \cite{Briegel etc}
  which is based on certain quantum states.  Their graphical notation
  is dual to ours, with the vertices representing qubits and the edges
  representing entanglement between the individual quibits.  Within
  the formalism of this paper, a state is simply a process with no
  inputs; that  is, a concrete graph with empty domain.  An important
  application is to show that states or computations implmented
  differently are equivalent:  this includes graph states, and
  measurement based computations more generally. 
\end{remark}
